How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-06-19 doc/paper.tex
paper has all figures
edits complete. pass to seth

\begin{abstract}
  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. We find that our model that includes our historical covariates (e.g. change in geographic range) and allows covariate effects to vary over time to be overfit to our data, yielding biased forecasts of future extinction events. In contrast, our models which include either historical covariates or which allow for covariate effects to vary over time yield equivalent forecasts. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspect of these results is that our results is that our model that includes historical covariates (e.g. change in geographic range) and allows for covariate effects to vary over time is overfit to our data and produces more biased out-of-sample predictions than the other three models (Fig. \ref{fig:auc_compare}). In comparison, the models that include either the historical covariates or allows covariate effects to vary over time produce less biased out-of-sample forecasts. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

We also find that our three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model based on in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode indicate that this model is overfit to our data and produces less accurate forecasts than the other three models \ref{fig:fold_auc}). Including historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to a model that allows covariate effects to vary over time but does not include historical covariates. This contrasts with previous findings using geological stages as their time unit \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.

It is interesting to note that both historical covariates like change in geographic range and allowing covariate effects to vary over time can be thought of as different ways of encoding the past: explicitly or implicitly. Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model. By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}).

@@

We present the group-level estimates are the average effects of that covariate over time and across taxonomic groups. These estimates make up the vector $\mu$ described above. In addition to the average effect of our covariates, we also included the estimate of the group-level intercept, which is the average log-odds of extinction for an average observation over time and across taxonomic groups.

\begin{figure}[ht]
  \centering
  \includegraphics{../results/figure/effect_est}
  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}). The posterior distribution of our parameter estimates are presented as densities. Below each of these densities is marked the median estimate along with 50\% and 80\% credible intervals. Estimates are on the log-odds scale.}
  \label{fig:param_est}
\end{figure}

Our VP model allows for these our regression coefficients to vary over time and between taxonomic groups. Here we present the population-level estimates for our regression coefficients along with the population-level intercept estimates. Population-level estimates are those for a specific time interval and taxonomic group. In this case, the intercept estimate describes the log-odds of extinction for an average observation at that point in time and taxonomic group.

\begin{figure}[<+htpb+>]
  \centering
  \includegraphics{../results/figure/eff_time_group}
  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}). Posterior estimates are presented as a time series. The black line represents the median estimate of that parameter. In addition, the 50\%, 80\%, and 95\% credible intervals are indicated. Estimates are on the log-odds scale.}
  \label{fig:param_est_time_group}
\end{figure}

>>>>

\begin{abstract}
  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. We find that models which include either historical covariates or account for variation in covariate effects over time yield equivalent forecasts, but a model including both is overfit and yields biased forecasts. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). 
One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig. \ref{fig:auc_compare}). However, we might expect a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

We find that three of our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time, the out-of-sample forecasts from this mode demonstrate this model yields biased forecasts, which indicate that this model is overfit to our data \ref{fig:fold_auc}). Including historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to a model that allows covariate effects to vary over time but does not include historical covariates.

It is interesting to note that both historical covariates like change in geographic range and allowing covariate effects to vary over time can be thought of as different ways of encoding the past: explicitly or implicitly. Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model. By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results supplement those of \citet{Kiessling2016} which examined using differences in geographic range over time to predict extinction risk.

It is imporatant to remember that our model with historical covariates assumes that extinction selection based on these covarites is constant over time. If human impacts substatially alter present and future extinction risk dynamics, the assumption the effects of covariates have not changed over time may be limit or bias our ability to predict extinction in truly novel environmental regimes. Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time. Thus, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.

@@

First, we present the group-level estimates of our regression coefficients and intercept (Fig. \ref{fig:param_est}). Group-level effects are the average effects of that covariate over time and across taxonomic groups. These estimates make up the vector $\mu$ described above. In addition to the average effect of our covariates, we also included the estimate of the group-level intercept, which is the average log-odds of extinction for an average observation over time and across taxonomic groups.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/effect_est}
  \caption{Group-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}). The posterior distribution of our parameter estimates are presented as densities. Below each of these densities is marked the median estimate along with 50\% and 80\% credible intervals. Estimates are on the log-odds scale.}
  \label{fig:param_est}
\end{figure}

Second, we present the population-level estimates for our regression coefficients along with the population-level intercept estimates (Fig. \ref{fig:param_est_time_group}. Our VP model allows for these our regression coefficients to vary over time and between taxonomic groups. Population-level estimates are those for a specific time interval and taxonomic group. In this case, the intercept estimate describes the log-odds of extinction for an average observation at that point in time and taxonomic group.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{../results/figure/eff_time_group}
  \caption{Population-level parameter posterior estimates from our VP model (Table \ref{tab:model_def}). Posterior estimates are presented as a time series. The black line represents the median estimate of that parameter. In addition, the 50\%, 80\%, and 95\% credible intervals are indicated. Estimates are on the log-odds scale.}
  \label{fig:param_est_time_group}
\end{figure}
